---
title: "Characterization and Classification of Diabetic Retinopathy"
author: "Santosh Gummidipundi"
date: "1/7/2020"
output:
  github_document:
    toc: true
    toc_depth: 2
    dev: png
---

```{r Set Options, echo = FALSE}
options(knitr.kable.NA = '') #Sets default value for NA values in kable package which outputs tables in HTML
formatSet = "markdown" #Sets the default format to html. This can be changed to latex if we're generating a PDF instead
seed <- 117 #Sets the seed prior to any and all random events for reproducability
```

```{r Optional Package install, echo = FALSE}
# install.packages(c("randomForest",
#                    "e1071",
#                    "foreign",
#                    "dplyr",
#                    "knitr",
#                    "kableExtra",
#                    "desctable",
#                    "devtools",
#                    "tidyr",
#                    "ggplot2",
#                    "caret",
#                    "caTools",
#                    "gbm",
#                    "klaR",
#                    "neuralnet",
#                    "xgboost",
#                    "randomForest"))
```

```{r Library Loads, echo = FALSE, warning = FALSE, message = FALSE}
library(foreign)
library(dplyr)
library(knitr)
library(kableExtra)
library(desctable)
library(devtools)
library(tidyr)
library(ggplot2)
library(caret)
library(caTools)
library(neuralnet)
```

```{r Custom Functions Load, echo = FALSE, message = FALSE}
#This pulls in some common functions I use for cleaning and describing data
source_url("https://raw.githubusercontent.com/sgummidipundi/sgummidipundi_r/master/clean_dataset.R")
source_url("https://raw.githubusercontent.com/sgummidipundi/sgummidipundi_r/master/describe.R")

yes_no <- function(x) {factor(x, levels = c(1,0), labels = c("Yes","No"))}
```



# Dataset Information

The data analyzed here has been sourced from the UCI Machine learning repository. https://archive.ics.uci.edu/ml/datasets/Diabetic+Retinopathy+Debrecen+Data+Set. 

"The dataset consists of features extracted from the Messidor image set to predict whether an image contains signs of diabetic retinopathy or not." - UCI Machine Learning Repository

## Data Dictionary
```{r DataDictionary Load, echo = TRUE, warning=FALSE}
#This pulls in another CSV which specifies the data dictionary
dd <- read.csv("data/data_dictionary.csv")

kable(dd, 
      format = formatSet,
      col.names = c("Col #","Variable","Description","Type"),
      booktabs = TRUE) %>%
  kable_styling(bootstrap_option = "striped",
                position = "center")
  
```


```{r Data Load, echo = TRUE}
#Read in data
df <- read.arff("data/messidor_features-1.arff")
```

```{r Data Processing, echo = TRUE}
#Set column names because there is no header in this data
colnames(df) <- c("quality",
                  "pre_screen",
                  "ma_1",
                  "ma_2",
                  "ma_3",
                  "ma_4",
                  "ma_5",
                  "ma_6",
                  "ex_1",
                  "ex_2",
                  "ex_3",
                  "ex_4",
                  "ex_5",
                  "ex_6",
                  "ex_7",
                  "ex_8",
                  "dist",
                  "diam",
                  "am_fm",
                  "class")

#Create a copy of the data for later use in model training and testing
df_copy <- df

#Change attributes for variables for use in descriptive analyses
df <- df %>%
      mutate_if(names(.) %in% c("quality","pre_screen","am_fm","class"), yes_no) 

#These variable dictionaries will be used in creating table 1
numeric_labels <- c("ma_1" = "alpha = 0.5",
                    "ma_2" = "alpha = 0.6",
                    "ma_3" = "alpha = 0.7",
                    "ma_4" = "alpha = 0.8",
                    "ma_5" = "alpha = 0.9",
                    "ma_6" = "alpha = 1.0",
                    "ex_1" = "alpha = 0.3",
                    "ex_2" = "alpha = 0.4",
                    "ex_3" = "alpha = 0.5",
                    "ex_4" = "alpha = 0.6",
                    "ex_5" = "alpha = 0.7",
                    "ex_6" = "alpha = 0.8",
                    "ex_7" = "alpha = 0.9",
                    "ex_8" = "alpha = 1.0",
                    "dist" = "Distance – Center of Macula",
                    "diam" = "Diameter – Optic Disc")

categorical_labels <- c("am_fm" = "AM/FM Classification",
                        "quality" = "Quality Assessment",
                        "pre_screen" = "Pre-Screening Result")

```

# Objective

The objective of this analysis is to describe and characterize the image features extracted from the Messidor image set in addition to performing an array of ML models to predict whether a patient has diabetic retinopathy.

# Descriptive Analyses

Ordinarily, a study of predictive nature a descriptive analyses should not be performed for on a whole dataset and rather just the training dataset. However, for the purposes of this demonstration of programming and reproducibility it will be performed.

## Table 1

```{r DA - Table 1, echo = TRUE, results = 'asis', warning = FALSE, message = FALSE}
#This creates a very basic table 1 plot for getting an overview of predictors and class balance.
#Summaries for overall group
summary_numeric <- df %>%
                   select_if(is.numeric) %>%
                   desctable(stats =  list("N" = length,
                                           "% or Mean (SD)" = is.numeric ~ mean_sd),
                   labels = numeric_labels) %>%
                   as.data.frame

summary_cat <- df %>%
               select(-class) %>%
               select_if(is.factor) %>%
               desctable(stats =  list("N" = length,
                                       "% or Mean (SD)" = is.factor ~ percent),
               labels = categorical_labels) %>%
               as.data.frame %>%
               mutate_if(is.numeric, funs(round(., digits = 1)))

summary <- rbind(summary_numeric, summary_cat) 
colnames(summary)[1] <- "var"


#Summaries stratified by class
summary_numeric_class <- df %>%
                         group_by(class) %>%
                         select_if(is.numeric) %>%
                         desctable(stats =  list("N" = length,
                                                 "% or Mean (SD)" = is.numeric ~ mean_sd),
                         labels = numeric_labels) %>%
                         as.data.frame

summary_cat_class <- df %>%
                     select_if(is.factor) %>%
                     group_by(class) %>%
                     desctable(stats =  list("N" = length,
                                             "% or Mean (SD)" = is.factor ~ percent),
                     labels = categorical_labels) %>%
                     as.data.frame %>%
                     mutate_if(is.numeric, funs(round(., digits = 1)))

summary_class <- rbind(summary_numeric_class, summary_cat_class)
colnames(summary_class)[1] <- "var"



summary <- cbind(summary, summary_class %>% select(-var))
summary$var <- gsub(".*:","",summary$var, perl = TRUE)


           
y <- kable(summary,
           format = "html",
           booktabs = TRUE,
           col.names = c(" ","N","%","N","%","N","%","p-value","test"),
           align = c("l",rep("c",8)),
           caption = "Table 1") %>%
    add_header_above(header = c(" " = 1, " " = 2, "DR Positive (n = 611)" = 2, "DR Negative (n = 540)" = 2,"Tests" = 2), align = "c") %>%
    add_header_above(header = c(" " = 1, "Overall" = 2, "Stratified" = 6), align = "c") %>%
    kable_styling(full_width = TRUE) %>%
    group_rows("Microaneurysms",1,6) %>%
    group_rows("Exudate Pixels",7,14) %>%
    group_rows("Quality Assessment",17,19) %>%
    group_rows("Pre-Screening Result",20,22) %>%
    group_rows("AM/FM Classification",23,25) %>%
    

print(y)
```

## Descriptive Plots

```{r DA Plots, echo = TRUE}
#This section details select graphical plots for variables of interest
#----------------------------Mean Exudates/ROI-------------------------------#
df_summary <- df %>%
              select(starts_with("ex"),"class") %>%
              group_by(class) %>%
              summarize_all(mean) %>%
              gather(ex, value, ex_1:ex_8)
df_summary$ex <- factor(df_summary$ex,
                        levels = c("ex_1",
                                   "ex_2",
                                   "ex_3",
                                   "ex_4",
                                   "ex_5",
                                   "ex_6",
                                   "ex_7",
                                   "ex_8"),
                        labels = seq(0.3,1.0,0.1))
              
plot <- ggplot(df_summary,
               aes(x = ex,
                   y = value,
                   fill = ex)) +
        geom_bar(position = "dodge",
                 stat = "identity",
                 aes(y = value,
                     x = factor(ex))) +
        theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
              legend.position = "none") +
        xlab("CI Alpha") +
        ylab("Mean Exudate/ROI") +
        ggtitle("Mean Exudate/ROI Stratified by Diabetic Retinopathy Outcome") +
  facet_grid(class ~ .)
plot

#----------------------------Mean Microaneurysms-------------------------------#
df_summary <- df %>%
              select(starts_with("ma"),"class") %>%
              group_by(class) %>%
              summarize_all(mean) %>%
              gather(ma, value, ma_1:ma_6)
df_summary$ma <- factor(df_summary$ma,
                        levels = c("ma_1",
                                   "ma_2",
                                   "ma_3",
                                   "ma_4",
                                   "ma_5",
                                   "ma_6"),
                        labels = seq(0.5,1.0,0.1))
              
plot <- ggplot(df_summary,
               aes(x = ma,
                   y = value,
                   fill = ma)) +
        geom_bar(position = "dodge",
                 stat = "identity",
                 aes(y = value,
                     x = factor(ma))) +
        theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
              legend.position = "none") +
        xlab("CI Alpha") +
        ylab("Mean Microaneurysms") +
        ggtitle("Mean Microaneurysms Stratified by Diabetic Retinopathy Outcome") +
  facet_grid(class ~ .)
plot

#-------------------AM/FM Classification---------------------------------#
df_summary <- df %>%
              select("am_fm","class") %>%
              mutate(am_fm = as.character(am_fm)) %>%
              group_by(class, am_fm) %>%
              tally %>%
              mutate(sum = sum(n)) %>%
              mutate(f = n/sum, am_fm = factor(am_fm, levels = c("Yes","No"), labels = c("AM","FM")))

plots <- ggplot(df_summary,
                aes(x = factor(am_fm),
                    weight = f,
                    fill = class)) +
         geom_bar(position = "dodge",
                  stat = "identity",
                  aes(y = f,
                      x = factor(am_fm))) +
        theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
              legend.position = "none") +
        facet_grid(class ~., scales = "free") +
        xlab("AM/FM Classification") +
        ylab("Proportion") +
        ggtitle("Proportion AM/FM Stratified by Diabetic Retinopathy Outcome") +
        scale_y_continuous(limits = c(0,1))
plots


#-------------------Pre-Screening Result---------------------------------#
df_summary <- df %>%
              select("pre_screen","class") %>%
              mutate(am_fm = as.character(pre_screen)) %>%
              group_by(class, pre_screen) %>%
              tally %>%
              mutate(sum = sum(n)) %>%
              mutate(f = n/sum, pre_screen = factor(pre_screen, levels = c("Yes","No"), labels = c("Severe Retinoabnormality","None")))

plots <- ggplot(df_summary,
                aes(x = factor(pre_screen),
                    weight = f,
                    fill = class)) +
         geom_bar(position = "dodge",
                  stat = "identity",
                  aes(y = f,
                      x = factor(pre_screen))) +
        theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
              legend.position = "none") +
        facet_grid(class ~., scales = "free") +
        xlab("Severe Retinoabnormality") +
        ylab("Proportion") +
        ggtitle("Proportion with Severe Retinoabnormality Stratified by Diabetic Retinopathy Outcome") +
        scale_y_continuous(limits = c(0,1))
plots

```

# Predictions

A train-test set is created by random shuffling and a 75/25 split.

```{r Split, echo = TRUE, message=FALSE}
set.seed(seed)
index <- sample(seq(1,nrow(df_copy)))
df_copy <- df_copy[index,]

df_copy$split <- sample.split(df_copy$ex_4, SplitRatio = 0.75)

train <- df_copy %>% 
         filter(split == TRUE) %>% 
         select(-split, -quality)

test <- df_copy %>% 
        filter(split == FALSE) %>% 
        select(-split, -quality)
```

## Feature Selection

```{r Feature Selection, echo = TRUE, message=FALSE, cache = TRUE}
set.seed(seed)
#Create an object to specify details of model selection
control <- rfeControl(functions = rfFuncs,
                      method = "repeatedcv",
                      number = 5,
                      repeats = 3)

results <- rfe(x = train[,c(1:18)],
               y = train[,19],
               sizes = c(1:18),
               rfeControl = control)
results

#Predictors to use for model training and testing
predictors <- c("ma_1", "ex_7", "ma_2", "ex_1", "ex_8")
```

From this we found the top 5 variables to be:

1. Microaneurysm, alpha = 0.5
2. Microaneurysm, alpha = 0.6
3. Exudates/ROI, alpha = 0.3
4. Exudates/ROI, alpha = 0.9
5. Exudates/ROI, alpha = 1.0

## Model Training

From here, we train several select models using the above predictors.

```{r Model Training, echo = TRUE, message=FALSE}
#Create a control object to specify training details
control <- trainControl(method = 'cv',
                        number = 3,
                        returnResamp = "none",
                        summaryFunction = twoClassSummary,
                        classProbs = TRUE)

```

### Logistic Regression

```{r Logistic Regression, echo=FALSE, warning=FALSE}
set.seed(seed)
#Fit Logistic Regression
lr <- train(train[,predictors], train[,"class"],
            method = "glm",
            family = "binomial",
            tuneLength = 10)

#Caclulate confusion matrix for predictions
confusionMatrix(data = predict(lr, test),
                reference = test$class)

```

### Random Forest
```{r RandomForest, echo = TRUE, warning=FALSE}
set.seed(seed)
#Train Random Forest
rf <- train(train[,predictors], factor(train[,"class"], levels = c(1,0), labels = c("yes","no")),
                 method = "rf",
                 trControl = control,
                 metric = "ROC",
                 preProc = c("center","scale"),
                 tuneLength = 3)

#Caclulate confusion matrix for predictions
confusionMatrix(data = predict(rf, test),
                reference = factor(test$class, levels = c(1,0), labels = c("yes","no")))

```

### Gradient Boosted Trees
```{r GBM - 1, echo = TRUE, message = FALSE, warning=FALSE}
set.seed(seed)
#Train gradient boosted machines
gbm <- train(train[,predictors], factor(train[,"class"], levels = c(1,0), labels = c("yes","no")),
             method = "gbm",
             trControl = control,
             metric = "ROC",
             preProc = c("center","scale"),
             tuneLength = 3,
             verbose = FALSE)
```

```{r GBM - 2, echo = TRUE, warning=FALSE}
#Caclulate confusion matrix for predictions
confusionMatrix(data = predict(gbm, test),
                reference = factor(test$class, levels = c(1,0), labels = c("yes","no")))
```


### k-Nearest Neighbors
```{r KNN, echo = TRUE, warning=FALSE}
set.seed(seed)
#Train KNN
knn <- train(train[,predictors], factor(train[,"class"], levels = c(1,0), labels = c("yes","no")),
             method = "knn",
             trControl = control,
             metric = "ROC",
             preProcess = c("center","scale"),
             tuneLength = 3)

#Caclulate confusion matrix for predictions
confusionMatrix(data = predict(knn, test),
                reference = factor(test$class, levels = c(1,0), labels = c("yes","no")))

```

### Naive Bayes
```{r NB - 1, echo = TRUE, message = FALSE, warning=FALSE}
set.seed(seed)
#Train Naive Bayes
nb <- train(train[,predictors], factor(train[,"class"], levels = c(1,0), labels = c("yes","no")),
            method = "nb",
            tuneLength = 3,
            preProcess = c("center","scale"),
            trControl = control,
            verbose = FALSE)
```

```{r NB - 2, echo = TRUE, warning=FALSE}
#Caclulate confusion matrix for predictions
confusionMatrix(data = predict(nb, test),
                reference = factor(test$class, levels = c(1,0), labels = c("yes","no")))

```


### Neural Network

```{r Neural Network - 1, echo = TRUE, message=FALSE, warning=FALSE}
set.seed(seed)
#Train Neural Network
anNet <- train(train[,predictors], factor(train[,"class"], levels = c(1,0), labels = c("yes","no")),
            method = "nnet",
            tuneLength = 3,
            preProcess = c("center","scale"),
            trControl = control,
            verbose = FALSE)
```

```{r Neural Network - 2, echo = TRUE, warning=FALSE}
#Caclulate confusion matrix for predictions
confusionMatrix(data = predict(anNet, test),
                reference = factor(test$class, levels = c(1,0), labels = c("yes","no")))
```


# Results

Based on these results, we find that the model best suited for accurate predictions of Diabetic Retinopathy is the Neural Network.
